Edge sign prediction in general graphs and connection with Correlation Clustering

### 2017-07-06 gtx_algorithm.tex
typos

\begin{lemma}
  \label{lemma:gtx_lessnodes}
  During the execution of the \gtx{} algorithm, for all $t \leq K-2$, $|V_{t+2}| \leq
  \frac{2}{3}|V_t|$.
\end{lemma}
\begin{marginfigure}
  \centering
  \includegraphics[width=\linewidth]{assets/raw/lots_of_singletons_bw.png}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_manysingles}
\end{marginfigure}
\begin{proof}
  Say that at iteration $t$ of \gtx{}, we have $n_t$ nodes and assume a fraction $\alpha$ of them
  are gonna form singleton stars. This means there will be at most $\frac{1-\alpha}{2}n_t$
  \emph{proper} stars, because such stars are made of at least two nodes. Now fix a constant
  $\lambda \in (0,1)$, for instance $\lambda = \nicefrac{9}{10}$. If $\alpha \leq \lambda$, $n_{t+1}
  \leq (\frac{1-\alpha}{2} + \alpha)n_t = \frac{1+\alpha}{2}n_t = \frac{19}{20}n_t$ and we get a
  constant rate reduction of the number of nodes. On the other hand, if $\alpha > \lambda$, we could
  be in the situation of the previous example and have $\alpha$ tends to $1$. Then we only have a
  few proper stars and by construction, each singleton star is connected to least one proper star.
  In the worst case, those connections are uniform so that each of these proper star has a degree
  $\frac{\alpha n_t}{\frac{1-\alpha}{2}n_t} = \frac{2\alpha}{1-\alpha}$ which goes to infinity when
  $\alpha$ goes to $1$ and is greater than $18$ when $\alpha$ is greater than $\nicefrac{9}{10}$.
  Even ignoring the fact that those proper stars are connected, it follows that in the next
  iteration, the number of node will be reduced by the degree of those proper stars. Setting in the
  worst case $n_{t+1}=\frac{1+\alpha}{2}n_t$, we thus have $$n_{t+2} \leq
  \frac{n_{t+1}}{\frac{2\alpha}{1-\alpha}} = \frac{1-\alpha^2}{4\alpha}n_t$$
  \begin{marginfigure}
    \centering
    \includegraphics[width=\linewidth]{assets/tmp-code/gtx_reduc.pdf}
    \caption{The fraction of $n_t$ nodes remaining in $n_{t+2}$ for the three cases described in the
    main text.}
    % \label{fig:gtx_manysingles}
  \end{marginfigure}
  This defines three ways to go from $n_t$ to $n_{t+2}$: if the first iteration has $\alpha>\lambda$,
  then we just seen that in the worst case we have at least $n_{t+2}\leq
  \frac{1-\lambda^2}{4\lambda}n_t$; if the first iteration has $\alpha < \lambda$, then either the
  second also has  $\alpha < \lambda$ (in which case $n_{t+2}\leq
  \left(\frac{1+\lambda}{2}\right)^2n_t$) or the second has $\alpha>\lambda$ (in which case
  $n_{t+2}\leq \frac{1+\lambda}{2}n_t$). We can optimize over $\lambda$ by finding its value such
  that $\frac{1-\lambda^2}{4\lambda}=\frac{1+\lambda}{2}$, yielding $\lambda = \frac{1}{3}$. Then,
  that in any event, $n_{t+2}\leq \frac{2}{3}n_t$ as claimed.
\end{proof}

From that we conclude that the number of iterations of the \gtx{} algorithm satisfies $K=O(\log
|V_0|)$. We would like to prove a similar result on the edges so that the overall complexity of the
\gtx{} algorithm would be $O(|V_0|\log |V_0|+|E_0|)$. However, this does not follow directly from the
reduction of number of nodes. Indeed we can have a situation where most of the edges of $E_t$ are
between the stars of $G_t$ and thus will remain in the next iteration. Still, we hope that
similarly to the previous proof, this high density of outer edges will lead to strong collapse in
the next iteration.

>>>>

\begin{lemma}
  \label{lemma:gtx_lessnodes}
  During the execution of the \gtx{} algorithm, for all $t \leq K-2$, $|V_{t+2}| \leq
  \frac{2}{3}|V_t|$.
\end{lemma}
\begin{marginfigure}
  \centering
  \includegraphics[width=\linewidth]{assets/raw/lots_of_singletons_bw.png}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_manysingles}
\end{marginfigure}
\begin{proof}
  Say that at iteration $t$ of \gtx{}, we have $n_t$ nodes and assume a fraction $\alpha$ of them
  will form singleton stars during the \tth{} iteration. This means there will be at most $\frac{1-\alpha}{2}n_t$
  \emph{proper} stars, because such stars are made of at least two nodes. Now fix a constant
  $\lambda \in (0,1)$, for instance $\lambda = \nicefrac{9}{10}$. If $\alpha \leq \lambda$, $n_{t+1}
  \leq (\frac{1-\alpha}{2} + \alpha)n_t = \frac{1+\alpha}{2}n_t$ and $\frac{1+\alpha}{2}n_t \leq
  \frac{1+\lambda}{2}n_t = \frac{19}{20}n_t$ so we get a
  constant rate reduction of the number of nodes. On the other hand, if $\alpha > \lambda$, we could
  be in the situation of the previous example and have $\alpha$ be arbitrarily close to $1$. Then we
  only have a few proper stars and by construction, each singleton star is connected to least one proper star.
  In the worst case, those connections are uniform so that each of these proper stars has degree at least
  $\frac{\alpha n_t}{\frac{1-\alpha}{2}n_t} = \frac{2\alpha}{1-\alpha}$ which goes to infinity when
  $\alpha$ goes to $1$ and is greater than $18$ when $\alpha$ is greater than $\nicefrac{9}{10}$.
  Even ignoring the fact that those proper stars are connected, it follows that in the next
  iteration, the number of node will be reduced by the degree of those proper stars. Setting in the
  worst case $n_{t+1}=\frac{1+\alpha}{2}n_t$, we thus have $$n_{t+2} \leq
  \frac{n_{t+1}}{\frac{2\alpha}{1-\alpha}} = \frac{1-\alpha^2}{4\alpha}n_t$$
  \begin{marginfigure}
    \centering
    \includegraphics[width=\linewidth]{assets/tmp-code/gtx_reduc.pdf}
    \caption{The fraction of $n_t$ nodes remaining in $n_{t+2}$ for the three cases described in the
    main text.}
    % \label{fig:gtx_manysingles}
  \end{marginfigure}
  This defines three ways to go from $n_t$ to $n_{t+2}$: if the first iteration has $\alpha>\lambda$,
  then we just seen that in the worst case we have at least $n_{t+2}\leq
  \frac{1-\lambda^2}{4\lambda}n_t$; if the first iteration has $\alpha < \lambda$, then either the
  second also has  $\alpha < \lambda$ (in which case $n_{t+2}\leq
  \left(\frac{1+\lambda}{2}\right)^2n_t$) or the second has $\alpha>\lambda$ (in which case
  $n_{t+2}\leq \frac{1+\lambda}{2}n_t$). We can optimize over $\lambda$ by finding its value such
  that $\frac{1-\lambda^2}{4\lambda}=\frac{1+\lambda}{2}$, yielding $\lambda = \frac{1}{3}$. Then,
  in any event, $n_{t+2}\leq \frac{2}{3}n_t$ as claimed.
\end{proof}

From that we conclude that the number of iterations of the \gtx{} algorithm satisfies $K=O(\log
|V_0|)$. We would like to prove a similar result on the edges so that the overall complexity of the
\gtx{} algorithm would be $O(|V_0|\log |V_0|+|E_0|)$. However, this does not follow directly from the
reduction of number of nodes. Indeed, we can have a situation where most of the edges of $E_t$ are
between the stars of $G_t$ and thus will remain in the next iteration. Still, we hope that
similarly to the previous proof, this high density of outer edges will lead to strong collapse in
the next iteration.
